---
title: 'DATA624: Homework 1'
author: "Donald Butler"
date: "2023-09-03"
output: 
  html_document:
    theme: yeti
    highlight: tango
    toc: true
    toc_float: true
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Homework 1

```{r loadLibraries, warning=FALSE, message=FALSE}
library(fpp3)
library(tidyverse)
library(ggplot2)
```

## Exercise 2.1

<blockquote>
Explore the following four time series: *Bricks* from *aus_production*, *Lynx* from *pelt*, *Close* from *gafa_stock*, *Demand* from *vic_elec*.  

* Use ? (or help()) to find out about the data in each series.
* What is the time interval of each series?
* Use autoplot() to produce a time plot of each series.
* For the last plot, modify the axis labels and title.
</blockquote>

### *Bricks* from *aus_production*

The time interval in the *aus_production* dataset is quarterly.

```{r}
#help(aus_production)
autoplot(aus_production, Bricks)
```

### *Lynx* from *pelt*

The time interval for the *pelt* dataset is annual.

```{r}
#help(pelt)
autoplot(pelt, Lynx)
```

### *Close* from *gafa_stock*

The time interval for the *gafa_stock* dataset is trading days of the stock market.

```{r}
help(gafa_stock)
autoplot(gafa_stock, Close)
```

### *Demand* from *vic_elec*

The time interval for the *vic_elec* dataset is every 30 minutes.

```{r}
help(vic_elec)
autoplot(vic_elec, Demand) + 
  ggtitle('Electricity Demand for Victoria, Australia', 
          subtitle = 'Half-Hour Intervals') + 
  labs(x = 'Time in 30 minute intervals', 
       y = 'Demand in MWh')
```

## Exercise 2.2

<blockquote>
Use *filter()* to find what days correspond to the peak closing price for each of the four stocks in *gafa_stock*.
</blockquote>

```{r}
gafa_stock %>%
  group_by(Symbol) %>%
  filter(Close == max(Close)) %>%
  select(Symbol,Date,Close)
```

## Exercise 2.3

<blockquote>
Download the file *tute1.csv* from [the book website]('https://bit.ly/fpptute1'), open it in Excel, and review its contents. You should find four columns of information. Columns B through D each contain a quarterly series, labelled Sales, AdBudget and GDP. Sales contains the quarterly sales for a small company over the period 1981-2005. AdBudget is the advertising budget and GDP is the gross domestic product. All series have been adjusted for inflation.
</blockquote>


## Exercise 2.4

<blockquote>
</blockquote>


## Exercise 2.5

<blockquote>
</blockquote>


## Exercise 2.8

<blockquote>
</blockquote>


